thesis2024

### 2025-05-06 01-intro.qmd
Project organised, all articles and chapters compilled and revised

The Iberian Peninsula has often been described as a refugium during the coldest periods of the Late Pleistocene [@gomezRefugiaRefugiaPatterns2007a; @gonzalez-samperiz_steppes_2010-1; @hewitt_genetic_2000; @jochimLatePleistoceneRefugia1987; @consuegraMitochondrialDNAVariation2002; @jonesMobilitySettlementResource2013; @bichoEarlyUpperPaleolithic2017], providing milder conditions than northern Europe and supporting stable human occupation throughout the UP. However, during abrupt climatic events such as Heinrich Events and the Last Glacial Maximum, the region experienced colder temperatures, reduced precipitation, and vegetation shifts. Even within Iberia, north/south differences have been identified: during abrupt climatic events, arboreal taxa such as oak were replaced by infertile heathlands in northern Iberia or semi-desert steppe vegetation in southern Iberia [@zapataPalinologicalRecordHeinrich2003; @fletcherMillennialscaleVariabilityLast2010; @fletcherOrbitalSuborbitalscaleClimate2008; @naughtonWetDryClimatic2009; @sanchezgoniEuropeanClimaticResponse2000; @sanchezgoniContrastingImpactsDansgaard2008; @naughtonPresentdayLast252007; @roucouxCombinedMarineProxy2001; @boessenkoolNorthAtlanticSeasurface2001; @turonLandseaCorrelationsLast2003; @roucouxResponseNWIberian2005]. This would have resulted in the expansion of dry, risk-prone environments. Instead of being uniformly hospitable, Iberia may have exhibited a "leopard coat" distribution of ecological niches [@schmidtRapidClimateChange2012; @jennings_southern_2011]. These niches likely supported continuous human occupation, leading to the formation of multicomponent sites and eco-cultural niches [@yravedraNotDesertedpaleoecologyHuman2016; @hawsPaleolithicSocionaturalRelationships2012; @cascalheira_2017; @cascalheiraHunterGathererEcodynamics2013; @banksHumanEcologicalNiches2008; @forteaperezCuevaMallaetesProblemas1975].

@@

Vale Boi, located in southwestern Iberia, is a multi-component site spanning several UP technocomplexes, from the Gravettian (c. 32 ka cal BP) to the Magdalenian [c. 15 ka cal BP; @bichoRockshelterStudiesSouthwestern2007; @bichoEdgeCaseVale2012; @bicho_ecodynamics_2013; @cascalheira_2017; @cascalheiraHunterGathererEcodynamics2013; @manneIntensiveSubsistencePractices2012; @mannePryingNewMeaning2011]. The site, considered an eco-cultural niche, provided stable climatic conditions even during colder periods, supporting continuous residential occupations that have been consistently dated [@bicho_ecodynamics_2013; @cascalheira_2017; @belmiro_gravettian-solutrean_2021]. Extensive lithic assemblages from Vale Boi have undergone technological analysis, revealing patterns of technological change and resilience [@belmiro_gravettian-solutrean_2021; @cascalheiraValeBoiAlgarve2013; @gibajabaoProvenienceTechnologyMorphology2013; @horta_role_2019; @marreirosLithicTechnologyGravettian2015]. The site has also been identified as a centre for knowledge transfer, particularly during the Last Glacial Maximum [@cascalheiraINFLUENCIAMEDITERRANICANAS2013; @cascalheiraTerritorialityOrganizationTechnology2019].

>>>>

The Iberian Peninsula has often been described as a refugium during the coldest periods of the Late Pleistocene [@gomezRefugiaRefugiaPatterns2007; @gonzalez-samperiz_steppes_2010-1; @hewitt_genetic_2000; @jochimLatePleistoceneRefugia1987; @consuegraMitochondrialDNAVariation2002; @jonesMobilitySettlementResource2013; @bichoEarlyUpperPaleolithic2017], providing milder conditions than northern Europe and supporting stable human occupation throughout the UP. However, during abrupt climatic events such as Heinrich Events and the Last Glacial Maximum, the region experienced colder temperatures, reduced precipitation, and vegetation shifts. Even within Iberia, north/south differences have been identified: during abrupt climatic events, arboreal taxa such as oak were replaced by infertile heathlands in northern Iberia or semi-desert steppe vegetation in southern Iberia [@zapataPalinologicalRecordHeinrich2003; @fletcherMillennialscaleVariabilityLast2010; @fletcherOrbitalSuborbitalscaleClimate2008; @naughtonWetDryClimatic2009; @sanchezgoniEuropeanClimaticResponse2000; @sanchezgoniContrastingImpactsDansgaard2008; @naughtonPresentdayLast252007; @roucouxCombinedMarineProxy2001; @boessenkoolNorthAtlanticSeasurface2001; @turonLandseaCorrelationsLast2003; @roucouxResponseNWIberian2005]. This would have resulted in the expansion of dry, risk-prone environments. Instead of being uniformly hospitable, Iberia may have exhibited a "leopard coat" distribution of ecological niches [@schmidtRapidClimateChange2012; @jennings_southern_2011]. These niches likely supported continuous human occupation, leading to the formation of multicomponent sites and eco-cultural niches [@yravedraNotDesertedpaleoecologyHuman2016; @hawsPaleolithicSocionaturalRelationships2012; @cascalheira_2017; @cascalheiraHunterGathererEcodynamics2013; @banksHumanEcologicalNiches2008; @forteaperezCuevaMallaetesProblemas1975].

@@

Vale Boi, located in southwestern Iberia, is a multi-component site spanning several UP technocomplexes, from the Gravettian (c. 32 ka cal BP) to the Magdalenian [c. 15 ka cal BP; @bichoRockshelterStudiesSouthwestern2007; @bichoEdgeCaseVale2012; @bicho_ecodynamics_2013; @cascalheira_2017; @cascalheiraHunterGathererEcodynamics2013; @manneIntensiveSubsistencePractices2012; @mannePryingNewMeaning2011]. The site, considered an eco-cultural niche, provided stable climatic conditions even during colder periods, supporting continuous residential occupations that have been consistently dated [@bicho_ecodynamics_2013; @cascalheira_2017; @belmiro_gravettian-solutrean_2021]. Extensive lithic assemblages from Vale Boi have undergone technological analysis, revealing patterns of technological change and resilience [@belmiro_gravettian-solutrean_2021; @cascalheiraValeBoiAlgarve2013; @gibajabaoProvenienceTechnologyMorphology2013; @horta_role_2019; @marreirosLithicTechnologyGravettian2015]. The site has also been identified as a centre for knowledge transfer, particularly during the Last Glacial Maximum [@cascalheiraINFLUENCIAMEDITERRANICANAS; @cascalheiraTerritorialityOrganizationTechnology2019].
